Master's Project Journal

### 2019-04-12 thesis/FPGABasedADPLLs.tex
just TDL left

\section{Phase Detector}
Three potential \acl{PD} designs will be addressed in this thesis, sign-only detecion with a Bang-Bang Detector, \ac{FPGA} clocked detection using a \ac{FSM} and an emulation of a \ac{TDL} using inverter primitives to create a delay, similarly to the \ac{RO} seen previously. Both designs will have to have a representation for both situations where the generated signal leads and lags the reference signal, and for the magnitude of this difference, a problem most obviously solved by using a signed binary representation of the difference. Most easily implemented in hardware is twos' complement, as operations can be carried out by the same hardware as unsigned arithmetic. The sign convention for phase error assigns the positive values to the situation where the generated signal lags the reference, and vice versa for the leading case. Apart from the Bang-Bang Detector, each design features two blocks, one to determine the sign of the error and a second which estimates the magnitude. These blocks could theoretically be mixed-and-matched, but as they operate on different paradigms no benefit would be obtained.

@@

\subsection{\acs{FPGA} Clocked Phase Detector}
The easier to implement of the two designs uses a combination of a state machine to determine which edge occurred first, and to control a twos' complement up-down counter the estimates the magnitude of the error.

\subsubsection{State Machine Based Detection}
\subsubsection{Up-Down Counter}
The up-down counter in this design performs estimates the magnitude of the phase difference between the reference and generated signals. The state machine provides control instructions, indicating whether the count should be in the increasing or decreasing directions, and and holds the counter in reset once the measurement interval has been completed. The counter counts in twos' complement, so counting in the downward direction can be carried out by adding on the representation of -1, the binary signal containing all 1s. The other advantage of counting in two's complement is that count value is already in the correct format for the follow block, the phase detector in a regular \ac{ADPLL} or the error combiner in this instance, therefore the count value at the end of the measurement interval can be stored in a register, which maintains the output value of the phase detector between measurements. Twos' complement is an advantageous way to represent signed numbers in hardware as the mathematical operations are identical to their unsigned counterparts, thus allowing use of any adder or carry hardware that may be present on the \ac{FPGA}.

@@

\subsection{SigNum Detector}
\subsubsection{Sign Detection}
\subsubsection{Tapped Delay Line}

@@

Again both integer and fixed-point methods of \ac{LF} design are possible, with the integer method requiring logic combining the two paths to preserve the relationship between the proportional and integral gains, which from the work of Koskin \textit{et al} is known to be several binary orders of magnitude \cite{koskin2018generation}. Failure to preserve a relationship of this order may lead to the filter becoming unstable, as in the upper-left region of Figure \ref{fig:gain_grid}. In the fixed-point system the relationship between $k_p$ and $k_i$ is preserved during the combination of proportional and integral paths as both values have their radix point aligned before addition is carried out.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.6\textwidth]{../eugene}%{../koskin_gain_grid} %TODO
	\caption[Stability of \acl{LF} gains]{Stability of \acl{LF} gains \cite{koskin2018generation}.}
	\label{fig:gain_grid}
\end{figure}

>>>>

\section{Phase Detector}
Three potential \acl{PD} designs will be addressed in this thesis, sign-only detection with a Bang-Bang Detector, \ac{FPGA} clocked detection using a \ac{FSM} and an emulation of a \ac{TDL} using inverter primitives to create a delay, similarly to the \ac{RO} seen previously. Both designs will have to have a representation for both situations where the generated signal leads and lags the reference signal, and for the magnitude of this difference, a problem most obviously solved by using a signed binary representation of the difference. Most easily implemented in hardware is two's complement, as operations can be carried out by the same hardware as unsigned arithmetic. The sign convention for phase error assigns the positive values to the situation where the generated signal lags the reference, and vice versa for the leading case. Apart from the Bang-Bang Detector, each design features two blocks, one to determine the sign of the error and a second which estimates the magnitude. These blocks could theoretically be mixed-and-matched, but as they operate on different paradigms no benefit would be obtained.

@@

\subsection{\acs{FPGA} Clocked Phase Detector}
The easier to implement of the two designs uses a combination of a state machine to determine which edge occurred first, and to control a two's complement up-down counter, which acts as the \ac{TDC} in this detector, and estimates the magnitude of the error.

\subsubsection{State Machine Based Detection}
While sign detection clocked by the \ac{FPGA} can be performed in a variety of ways, a state machine is a straightforward way in which to accomplish this. A \acl{FSM} consists of three main components: the states themselves, the next state logic and the output logic. In this case the next state logic is the generated and reference signals, and the output logic sets the count instructions for the up-down counter as well as ensuring the phase error output is held constant between measurement periods. The \ac{FSM} will have a state representing the case where either the reference or generated signals occurred first, in addition to a wait state used between the end of one measurement interval and the commencement of the next. Two types of \ac{FSM} exist, Moore and Mealy Machines, the difference being that Moore Machines base their output only on the current state while Mealy Machines do so on both the current state and the inputs to the system. If a Moore Machines is used an extra state is required in order to hold the value of the phase error at the end of the measurement interval.

As this design is clocked by the \ac{FPGA}, the phase error is quantised to the the \ac{FPGA} clock periods, thus setting the resolution of the design. While, depending on the oscillator design, the generated clock may be synchronous to the \ac{FPGA}, the reference signal arrives from off-chip and thus will not be synchronised to the measurement clock. In order to avoid any metastability events arising from an input signal changing within the set-up and hold times of the registers sampling them, synchroniser circuits are used to ensure the value sampled is either logic high or low at the instance of the measurement. As these synchronisers are the first \ac{FPGA} clocked elements in the measurement chain, they also act as the system's quantisers. There is no degradation in detection resolution, as the signals would be quantised to the \ac{FPGA} clock anyway.

There are three situations that may occur while the \ac{FSM} is in the waiting state. Firstly the case where the reference signal sees a rising edge before the generated clock, and thus the generated clock is said to be lagging the reference. As previously mentioned the convention dictates that the phase difference have a positive sign, so the counter is incremented to count in the positive direction. Similarly if the generated signal is leading the reference the sign of the error is negative and thus the counter is instructed to count downwards. The third potential scenario occurs when both clock signals have a rising edge within the same period of the \ac{FPGA} clock, which due to quantisation will be interpreted as both signals occurring at exactly the same time. In this case the counter receives no instruction and a zero value is stored until the next measurement interval. Once the \ac{FSM} is in a measurement state, and the counter either incrementing or decrementing, a rising edge occurring on the opposite signal to that which initiated the measurement will end the measurement interval.

%TODO state trans diagram

\subsubsection{Up-Down Counter}
The up-down counter in this design performs estimates the magnitude of the phase difference between the reference and generated signals. The state machine provides control instructions, indicating whether the count should be in the increasing or decreasing directions, and and holds the counter in reset once the measurement interval has been completed. The counter counts in two's complement, so counting in the downward direction can be carried out by adding on the representation of -1, the binary signal containing all 1s. The other advantage of counting in two's complement is that count value is already in the correct format for the follow block, the phase detector in a regular \ac{ADPLL} or the error combiner in this instance, therefore the count value at the end of the measurement interval can be stored in a register, which maintains the output value of the phase detector between measurements. Two's complement is an advantageous way to represent signed numbers in hardware as the mathematical operations are identical to their unsigned counterparts, thus allowing use of any adder or carry hardware that may be present on the \ac{FPGA}. An important attribute of this counter is the ability to freeze the count at the minimum and maximum count values in order to avoid a situation where the counter overflows, thus at high phase offsets potentially alternating between positive and negative differences which may restrict the locking range of the \ac{ADPLL}. Disallowing overflow also aligns the behaviour of this phase detector with that of those implementing delay lines, allowing better interchangeability and comparison, which for a testing platform may be valuable.

@@

\subsection{SigNum Detector}
The SigNum detector is based on the methodology used in the Bang-Bang \acl{PD} seen previously, but rather than just performing sign detection, the magnitude of the phase difference can also be measured. In the case of the \ac{FPGA} clocked design the counter handled the representation of both sign and magnitude, but as the name SigNum (Sign Number) suggested both are handled separately in this design. The modified Bang-Bang detector establishes which signal occurred first and generates a control signal that is analogous to the time delay between a rising edge on each input signal. A \ac{TDL} then uses this control signal to convert the phase difference between the signals to a digital signal. This magnitude is combined with the sign detected by the bang-Bang detector to form a two's complement error signal.

This method of phase detection is advantageous as similarly to the \acl{RO}, the \ac{TDL} can be implemented using inverters, and therefore is not synchronised to the \ac{FPGA} clock and the propagation time through inverters sets the resolution, thus allowing a smaller step than that of the clocked designs. The disadvantages, however, are also consistent with the \ac{RO} in that the propagation time will vary significantly depending on the \ac{EDA}'s placement of the circuit. This is not as severe a limitation for the \ac{TDL} as the \ac{RO} as while the tuning range is affected by this variation it cannot cause the centre point to vary, whereas the impact of placement on the \ac{RO} caused both variation in range and centre frequency.

\subsubsection{Sign Detection}
The Bang-Bang detector structure changes significantly in order to provide the control signals required for the \ac{TDC}. The \ac{TDC} expects a long pulse of logic high for duration of the measurement interval, going low once the interval has ended. The original single flip-flop design maintains a value between rising edges of the reference clock and could not generate the required pulse. Instead two flip-flops are used, each clocked on an input signal with their $D$ inputs connected to logic high. The $Q$ outputs of these registers are connected to an \acs{XOR} gate which generates a pulse as long as a rising edge has only been seen on one signal. Once a second rising edge occurs the \acs{XOR} output will be set to 0, thus terminating the pulse. On completion of the measurement interval, the flip-flops must be reset in preparation for the next measurement cycle, and this can be accomplished by ANDing together the $Q$ outputs of each signal, producing a \texttt{clear} signal that will go high when both flip-flops have a logic high output.

To determine the sign itself, which clock edge first had a rising edge must be recorded, and this can be done by setting a \ac{SR} latch if the rising edge of the generated signal occurs first and resetting in the case of the reference clock having the first rising edge. Looking at the whole measurement cycle however, it is apparent that all 4 combinations possible of two inputs occur, putting the \ac{SR} latch into which the output is undefined. Careful design of the system can avoid problems arising from this. Examining the truth table in Figure \ref{fig:sr} (a), the combination to be avoided is the presence of a 1 on both outputs, precisely the situation seen at the end of the measurement interval. In order to avoid the undefined state occurring at the end of the measurement interval, both inputs and outputs to the \ac{SR} latch can be inverted using a NOT gate. Figure \ref{fig:sr} (b) illustrates how the $\overline{Q}$ output now behaves as $Q$ did previously, however the end of measurement state where both input registers have been triggered now causes the \ac{SR} latch to enter the hold state where it maintains the previous values. The sign, $\overline{Q}$ of the \ac{SR}, can then be loaded into a register using the \texttt{clear} signal generated previously, to maintain it until the next measurement completes.

This, however, ignores the potential for metastability which may arise if both signals have a rising edge within an extremely short period of time. While it may in other systems be feasible to dismiss this occurrence as a low probability event, the goal of a \ac{PLL} is to reduce the time delay between rising edges to zero which increases the possibility of this occurring, which may lead to indeterminate behaviour in other parts of the feedback network. In order to avoid this in an \ac{ASIC}, an arbitration circuit is placed between the \ac{SR} latch and the register storing the sign between measurements, which will default to one of the two signals in this case. This is however a mixed signal circuit, such as that proposed by Tierno \textit{et al} (2008) and implemented by Zianbetov and Shan \cite{tierno2008wide,zianbetov2013phd,shan2014phd}, which requires two \ac{PMOS} and two \ac{NMOS} transistors and cannot be implemented on an \ac{FPGA}. As the output of the \ac{SR} latch will be indeterminate prior to the detection of the first edge, the impact of both edge detection registers' output is that the \ac{SR} latch attempts to hold this indeterminate value as an output. To avoid this unwanted behaviour the arbiter present on an \ac{ASIC} is replaced by a second \ac{SR} latch, connected as in Figure \ref{fig:arbitration}. This ensures the value presented to the register storing the sign is always determinate.
%TODO sr truth table
%TODO image of simplified detector

\subsubsection{Tapped Delay Line}
To complement the asynchronous sign detection circuit, an asynchronous \ac{TDC} is also required, as otherwise there would be no benefit to using the above method over the more easily understood \ac{FSM} based design as quantisation to the clock period would happen in the \ac{TDC} rather than the sign detection circuitry. The standard method for implementing a \ac{TDC} on an \ac{ASIC} is through a tapped delay line, where a signal is propagated through a chain of fixed length delays, and at the end of the measurement interval it is recorded whether or not the signal has propagated to each point in the chain. As with the mixed signal circuitry previously addressed, a \ac{TDL} cannot be exactly replicated on an \ac{FPGA}, but through the use of inverters it is possible to replace the unrealisable fixed delays.

%TODO high level TDL image (using Z-1s)
%TODO eugene re: why no zero output

@@

Again both integer and fixed-point methods of \ac{LF} design are possible, with the integer method requiring logic combining the two paths to preserve the relationship between the proportional and integral gains, which from the work of Koskin \textit{et al} is known to be several binary orders of magnitude \cite{koskin2018generation}. Failure to preserve a relationship of this order may lead to the filter becoming unstable, as in the upper-left region of Figure \ref{fig:gain_grid}. In the fixed-point system the relationship between $k_p$ and $k_i$ is preserved during the combination of proportional and integral paths as both values have their radix point aligned before addition is carried out.
\begin{figure}[h]
	\centering
	\includegraphics[width=0.6\textwidth]{../eugene}
	\caption[Stability of \acl{LF} gains]{Stability of \acl{LF} gains \cite{koskin2018generation}.}
	\label{fig:gain_grid}
\end{figure}
